Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-15 response_to_referees.tex

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1\vskip 5mm}
\newcommand{\refb}[1]{\textbf{R2:} #1\vskip 5mm}
\newcommand{\editor}[1]{\textbf{Editor:} #1\vskip 5mm}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}

@@

\refa{1.2 Despite the more general formulation in the beginning, this
approach makes use only of the last time point in the trajectory,
rather than the full trace (contrary to the work of Illingworth\&
Mustonen). I find the observed neutral-like statistics not surprising
for such a limited data input, which doesn't capture past frequency dynamics. Therefore, these conclusions should be
revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}

\response{
	Illingworth and Mustonen (I\&M) presented a careful analysis to what extent trajectories can be fit using a model assuming a single fitness coefficient, or a model that differentiates different background fitnesses.
	Our results are consistent with their finding that a single parameter model (which only allows monotonic trajectories) does capture the majority of trajectories.
	Beyond that, our approach and aim differ from I\&M.
	We investigate a priori features of mutations that have information on whether the mutation fixes, not whether the trajectories can be parameterized by a model.
	Fitting a model to a partial trajectory to predict fitness is not expected to work: I\&M showed that subsequent interference has a substantial effect and our results indicate that the speed at which the mutation rises does not predict fitness.

	Furthermore, I\&M estimate fixing mutations to have a much higher intrinisic fitness than dying mutation while we show that very little such information is available a priori.
	The retrospective fit to trajectories tends to assign a positive fitness effect to fixing (and hence on average rising) mutations.
	This is why we resort to a model free approach ask what a priori information can help to determine which mutations fix or die.
}

% \responsedraft{
% 	We thank reviewer 1 for pointing out the work of Illingworth\&Mustonen to us, which we were unaware of. However, we believe that the method used in their article is not applicable as is in our case. To explain our thinking, we first briefly summarize the approach taken in Illingworth\&Mustonen: the authors fit two kind of models to fully known frequency trajectories:
% 	\begin{itemize}
% 		\item An "unlinked" model, where the fitness effect of a mutation fully determines its ultimate fixation or loss. The only trajectories that the model can produce are upward or downward sweeps, for beneficial or deleterious mutations respectively.
% 		\item A "linked" model, in which the fate of a mutation is determined both by its own fitness and by the fitness of other mutations present on the genomes it appears on. This model fits actual trajectories much more accurately, which is one of the main findings of the paper.
% 	\end{itemize}
% 	Interestingly, both models depend on the same parameters that have to be fitted, namely the fitness effects of individual mutations. Linkage between mutations in the second model is introduced by using the measured frequencies of appearance of two mutations on the same genome. \\
% 	\pierre{What I'm trying to say below
% 	\begin{itemize}
% 		\item Illingworth\&Mustonen don't do prediction, so there is no clear way to apply their method to our article.
% 		\item Fitting the initial part of the trajectory using their models doesn't make much sense.
% 		\item We want to stay model free: it's possible to find signs of selection without fitting (cf. H1N1/simulated data)
% 		\item Even if we're doing something very simple, it's surprising to see neutral-like behaviour.
% 	\end{itemize}}
% 	We first want to point out that this approach is fundamentally different to the one of our article. Since Illingworth\&Mustonen fit models to fully known trajectories, these models cannot meaningfully be used for prediction purposes. Indeed, fitness effects of individual mutations are fitted with knowledge of the fixation or loss of the given mutation, and of the frequency of other mutations in the case of the second model. This strongly contrasts with our approach which is to find patterns of predictability of the future behaviour of frequency trajectories using past data only: \emph{e.g.} if a frequency trajectory has risen from $0$ to a frequency $f>0$, what can be said of its future shape or of its fixation? Thus, methods in Illingworth\&Mustonen are designed to answer a different question than the one we are asking in our work. \\
% 	A way of reconciling the two approaches would be for us to fit the initial part of a trajectory with either model, and to use this fit to predict its future behaviour. However, we feel that this idea is not very relevant to our article for two reasons. First, neither model proposed by Illingworth\&Mustonen is really practical in our setting. The "unlinked" model only generates sweep-like trajectories, which our results show to be not representative of the "typical" trajectory (in good agreement with Illingworth\&Mustonen). The "linked" model can fit more complex trajectories, \emph{e.g.} ones that rise rapidly but ultimately die. However, it crucially depends on the frequencies of joint appearance of pairs mutations on genomes to introduce linkage. These frequencies vary through time, allowing the \emph{effective} fitness of a mutation to vary, and a trajectory to rise and then fall. In our case, these frequencies would only be known for the past, making impossible any prediction of the model. \\
% 	The second reason is that we believe that our model-free approach has intrinsic benefits. [Something saying that we'd prefer not to start fitting models to  data, because it's not the point. But I'm not sure how to formulate this.] \\
% 	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future. However, observing apparently-neutral statistics in the case of A/H3N2 is still surprising to us. The case of simulated populations shows that clear signs of selection can be observed by looking at basic features of trajectories (\emph{e.g.} figure S20 of the supplementary material). This is consistent with the theory of evolving population: if the only available information about a mutation is that it has risen in frequency starting from $0$, it has a higher chance on being beneficial than deleterious, and its fixation probability should be higher than what it would be in a neutral scenario. Hence, even with our very simple approach, we observe that predictability of A/H3N2 qualitatively differs from what models of adapting populations would suggest. \\}

@@

\refb{3. Question: what do these results mean for vaccines? How to decide which strains to use for vaccines? This may be known to those who work on Influenza, but for a relative outsider it is not clear. }

\response{
	Computational prediction of which circulating strains will dominate future influenza populations have been attempted for about 2 decades and have recently become a regular input to the biannual consultations on the influenza vaccine composition.
	Since performance of predictive models can only be assessed on limited historical data, we believe that our model-free approach clarifies which aspect of influenza evolution are predictable and which ones are not.

}

\refb{4. Question: at what point is it expected that the vaccine itself will influence frequency trajectories? See Wen et al Biorxiv 2020}
\response{
	The degree to which the vaccine drives evolution is a very interesting question!
	In the past, global influenza vaccine uptake was unlikely sufficient to generate much pressure and fairly thorough mixing of variants across countries with very different vaccination rates have been used to argue against a strong effect. But this might have changed in recent years and the paper by Wen et al in Viruses, 2018 picks up some signal.
	Overall, the case for strong effect is still weak to our knowledge.
}

\refb{Fig 1B is very hard to read. Maybe it should get more space (fig 1C could work with less space).}
\response{We have reworked this figure to allow more space for panel B.}

### 2020-12-15 article.tex
Typo

	A mutation with a fitness benefit should have a tendency to fix, while detrimental mutations fix less often.
	Fixation and loss of a mutation is, however, not only determined by its intrinsic fitness benefit, but also by the fitness of the background it arose on and the fitness of competing variants \citep{strelkowa_clonal_2012}.
	This degree to which these different components affect influenza evolution has been investigated by \citet{illingworth_components_2012}, who showed that intrinsic selection, background fitness, and later interference have comparable effects on fixation.
	These interference effects, along with stochastic epidemiological dynamics and transient selection, limit the predictability of evolution.
	Here, we investigate the frequency trajectories of amino acid mutations and the degree to which fixation or loss are predictable.
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not.
	To our surprise, we find that the predictability of these trajectories is very limited: For A/H3N2, the probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions.
	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability.
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

In supplementary figure \ref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for several predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020} including models that use serological data.
While many of these model predict future population composition, the only score that was predictive of fixation is mutations load.

>>>>

	A mutation with a fitness benefit should have a tendency to fix, while detrimental mutations fix less often.
	Fixation and loss of a mutation is, however, not only determined by its intrinsic fitness benefit, but also by the fitness of the background it arose on and the fitness of competing variants \citep{strelkowa_clonal_2012}.
	This degree to which these different components affect influenza evolution has been investigated by \citet{illingworth_components_2012}, who showed that intrinsic selection, background fitness, and later interference have comparable effects on fixation.
	These interference effects, along with stochastic epidemiological dynamics and transient selection, limit the predictability of evolution.
	Here, we investigate the frequency trajectories of amino acid mutations and the degree to which fixation or loss are predictable.
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not.
	To our surprise, we find that the predictability of these trajectories is very limited: for A/H3N2, the probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions.
	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability.
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

In supplementary figure \sref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for several predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020} including models that use serological data.
While many of these model predict future population composition, the only score that was predictive of fixation is mutations load.
